---
title: "p8105_hw3_mh4589"
author: "My An Huynh"
date: "2024-10-04"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)
data("ny_noaa")
library(ggridges)
```

The dataset NY NOAA provies information on temperature, snowfall, snow depth and precipitation everyday starting from 1981 at different weather stations in NY. 

There are `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. 
The most common snowfall is 0. This is because snow is commonly observed on certain days during winter months, but not the rest of the year. The next most commonly observed value is NA, which indicates missing data. 
```{r}
ny_noaa_df <- ny_noaa |> 
  separate(date, into = c("year", "month", "day"), sep = "-")

snow_mode =
  ny_noaa_df |> 
  count(snow, name = "frequency") |> 
  arrange(desc(frequency)) |> 
  slice(5)
```


Make a two-panel plot showing the average max temperature in January and in July in each station across years. I divided the temperatures by 10 to convert the units of measurements to degrees Celsius. The mean temperature in January is much lower than the mean temperature in July across all stations and across all years. There are a few outliers. 

```{r}
ny_noaa_df |> 
  drop_na(tmax, tmin) |> 
  filter(month == c("01", "07")) |>
  group_by(id, year) |> 
  mutate(
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin),
    tmax = tmax/10,
    tmin = tmin/10,
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = id, y = mean_tmax, color = year)) + 
  geom_point(aes(group = year), alpha = 0.5) +
  facet_grid(~month) +
  labs(
    title = "Mean Maximum Temperature Across Stations in January and July",
    x = "Station",
    y = "Mean Maximum Temperature (C)"
) +
  theme_minimal()
```

Hex plot showing tmax vs tmin 
```{r}
ny_noaa_df |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```


Disrtibution of snow by year 
```{r}
ny_noaa_df |> 
  filter(snow > 0, snow < 100) |> 
  ggplot(aes(x = snow, y = factor(year))) +
  geom_density_ridges() +
  labs(
    title = "Disribution of snow by year"
  )
  
```

## Problem 2

Loaded, tidied, merged and organized the acceleration and demographic datasets. Changed education from "1" to less than high school, "2" to high school equivalent, and "3" to more than high school. Changed sex from "1" to male and "2" to female. Encoded sex and education as factor variables. Excluded participants less than 21 years of age using filter, and those with missing demographic data using drop.na.

```{r}
accel_df =
  read_csv("data/nhanes_accel.csv") |> 
  janitor::clean_names()

covar_df = 
  read_csv("data/nhanes_covar.csv", skip = 4) |> 
  janitor::clean_names() |> 
  mutate(
    sex = recode(sex, 
                 "1" = "male", 
                 "2" = "female"),
    sex = factor(sex, levels = c("male", "female")),
    education = recode(education,
                       "1" = "Less than high school",
                       "2" = "High school equivalent",
                       "3" = "More than high school"),
    education = factor(education, levels = c("Less than high school", "High school equivalent", "More than high school"))
  )

nhanes_df = 
  left_join(covar_df, accel_df, by = "seqn") |> 
  filter(age >= 21) |> 
  drop_na(education, bmi, age, sex)

  
```

Created a reader-friendly table using `kable()` showing the number of men and women in each education category, and created a density plot showing the distribution of age for men and women across each education category. 

The table shows that the "more than high school" group had the most participants (total of 115), while the "less than high school" and "high school equivalent" groups have 55 and 58, respectively. The ratio between male and female participants for "less than high school" and "more than high school" groups were approximately 1:1, whereas the "high school equivalent" group have a male-to-female ratio of 1.5:1. These differences may influence the other observed variables, such as total activity.

The density plot gives more information on the distribution of participants' age by sex and education.

For the "less than high school" group, the distribution for male participants is bimodal, with two peaks around 40-50 and 70-80.  The distribution of female participants is relatively unimodal, with a peak around ages 70-80. The peaks for both male and female at 70-80 suggest that there are more participants in this age range for this education category. There are approximately the same numbers of male and female participants aged 20-30, and the overall density is lower compared to other age groups. There are more male participants than there are female within the age range 30-45, and vice versa for the age range 45 to 80. 

For the "high school equivalent" group, the distribution for male participants is bimodal, with two peaks around ages 30-40 and 60-70. The distribution for female participants is unimodal and skewed towards older age ranges (60 or above). There are noticeably fewer female than male participants within the age range 20-40, and vice versa for the age range 70-80. The overall distirbution of age is relatively similar to that of the "less than high school" group. 

For the "more than high school" group, the distribution skews towards younger ages compared to the other two education groups. There are higher densities of both female and male participants within the age range 20-40. The density drops more sharply for older ages (60 or above) compared to the other two education groups.

```{r}
nhanes_df |> 
  group_by(education, sex) |> 
  summarize(
    count = n()
  ) |> 
  pivot_wider(
    names_from = education,
    values_from = count
  ) |> 
  knitr::kable(
    caption = "Number of men and women in each education category"
  ) 

nhanes_df |> 
  ggplot(aes(x = age, fill = sex)) + 
  geom_density(alpha = 0.5) +
  facet_grid(. ~ education) +
  labs(
    title = "Distribution of Age by Sex and Education",
    x = "Age",
    y = "Density"
  ) 
  
  theme_minimal()
```

Created a new `total_activity` variable that aggregates activity level across 1440 minutes for each participant. Made a scatter plot of total activity against age and included a trend line.

The overall trend indicates that activity levels decrease as age increases for all participants, male and female and across 3 education groups. 

For the "less than high school" group, total activity is highest for ages around 20 and decline as age increases for both men and women. There is a sharp increase in activity among male participants around the age range 55-65, followed by a sharp decrease. There is a slight increase in activity among female participants around 55-65 as well, but this change is not as sharp as was observed in the male group, and followed by a sharp decrease as age increases to 80. 

Similar to the distribution of the "less than high school" group, there is an overall decline in total activity as age increases for the "high school equivalent" group, but the decline is not as sharp. The distribution for total activity among female participants is bimodal, with two peaks indicating increased activity levels among ages 40-50 (denoted by larger peak) and 70-80 (denoted by smaller peak). The distribution for total activity among male participants only has only peak that corresponds to the age range 30-40. Total activity stays constant for male participants aged 60-80. The activity level among females is overall higher than that among males.

For the "more than high school" group, total activity stays relatively constant among female participants aged 20-60, followed by a sharp decrease for ages 60 to 80. The total activity among male participants peaks around ages 40-50, and decreases for ages 50 to 80. Overall, total activity among female participants in this group is higher than that among male participants. 

```{r, warning = FALSE}
nhanes_df =
  nhanes_df |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minutes",
    values_to = "mims",
    names_prefix = "min"
  ) 

nhanes_df |> 
  group_by(seqn, sex, age, education) |> 
  summarize(
    total_activity = sum(mims)
  ) |> 
  ggplot(aes(x = age, y = total_activity, color = sex))+
  geom_point(alpha = 0.5)+
  geom_smooth(se = FALSE) +
  facet_grid(~education) + 
  labs(
    title = "Total Activity vs Age across Sex and Education Level",
    x = "Age (years)",
    y = "Total Activity over 24 hours"
  )

```

Made a three-panel plot that shows the 24-hour activity time courses for each education level and used color to indicate sex.

The trend in 24-hour activity is overall very similar among three education groups, with not much variability between males and females; there is higher activity around 12 am - 6 am, followed by a dip in activity levels at around 6 am to 3 pm indicating potential rest time, and activity picks up again from 3 pm to 12 am. The smoothed trend lines for males and females are quite similar, indicating no significant differences in the timing of activity throughout the day across the two sexes and education levels.

For the "less than high school" group, the dip is not as pronounced for males from around 9 am to 12 pm compared to that of females. This difference is not as noticeable for the "high school equivalent" group. 

Compared to the "less than high school" and "high school equivalent" groups, the "more than high school" group shows more variability in the patterns. Even though the smoothed lines look similar to the other two groups, the data itself shows noticeable sharp peaks around 3 am to 5 am and smaller sharp peaks from 1 pm to 8 pm for males. This suggests that some male participants may have varying workout schedules earlier or later during the day. There is a sharp peak in activity around 12 pm to 3 pm among females, indicating that females are more active around this time period. The overall variability indicates that participants in this group may have more diverse daily schedules or engage in a wider range of activities throughout the day. This variability may also be due to the fact that there is double the number of participants for this group than the other two groups, which increases variation. 

In conclusion, we may extrapolate that participants with more than a high school degree engage in varying activities throughout the day, with potential factors being flexible work hours and schedules, and different lifestyle habits. The dip in activity level around mid day may reflect lunch breaks or rest periods and is consistent among three education groups.  

```{r}
nhanes_df |> 
  ggplot(aes(x = minutes, y = mims, color = sex)) + 
  geom_line(alpha = .25) +
  facet_grid(.~ education) +
  geom_smooth(aes(group = sex), se = FALSE) + 
  labs(
    title = "24-hour Activity Time Courses for Each Education Level",
    x = "Time Course over 24 hours (Minutes)",
    y = "Activity"
  )
  
```

## Problem 3

Loaded the zip files and joined 4 datasets together. 

The joined dataset includes information on 1% of all Citi Bike rides in NYC with a total duration less than 4 hours in each of four months (January 2020, July 2020, January 2024 and July 2024). There are 9 variables and 99485 observations. The important variables in this dataset are ride IDs, types of ride (classic bike or electric bike), days of the week that the bikes were rented on (Monday through Sunday), duration of rental (in minutes), pickup and drop off locations, and types of membership (Citi Bike member or casual). After checking for NA, there are 232 rows with at least 1 NA value, and most of the NA values are within the `end_station_name` column. Since this does not seem to affect the next steps of data processing, I did not remove these NA values. 

```{r}
jan_20_df = 
  read_csv("data/Jan 2020 Citi.csv") |>
  janitor::clean_names() |> 
  mutate(
    month = "January",
    year = "2020"
  )

jan_24_df = 
  read_csv("data/Jan 2024 Citi.csv") |> 
  janitor::clean_names() |> 
   mutate(
    month = "January",
    year = "2024"
  )


jul_20_df = 
  read_csv("data/July 2020 Citi.csv") |> 
  janitor::clean_names() |> 
   mutate(
    month = "July",
    year = "2020"
  )


jul_24_df = 
  read_csv("data/July 2024 Citi.csv") |> 
  janitor::clean_names() |> 
   mutate(
    month = "July",
    year = "2024"
  )

citi_df = 
  bind_rows(jan_20_df, jan_24_df, jul_20_df, jul_24_df)
  
citi_df[!complete.cases(citi_df), ]

```

Created a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members. 

The results show that there are more bikes being rented in July compared to January across both 2020 and 2024, regardless of membership type. There are more bikes being rented in 2024 compared to 2020. Overall, there are more members renting CitiBike than non-members (casual). 

```{r}
citi_df |> 
  group_by(year, month, member_casual) |> 
  summarize(
    count = n()
  ) |> 
  pivot_wider(
    names_from = member_casual,
    values_from = count
  ) |> 
  knitr::kable()
```

Crete a table showing 5 most popular starting stations in July 2024, including the number of rides for each station. 

```{r}

 citi_df |> 
  filter(month == "July", year == "2024", .keep_all = TRUE) |> 
  group_by(start_station_name) |> 
  summarize(numer_of_rides = n()) |> 
  mutate(
    start_station_rank = min_rank(desc(numer_of_rides))
  ) |> 
  arrange(start_station_rank) |>
  select(-start_station_rank) |> 
  head(5) |> 
  knitr::kable()
```


Made a plot that investigates the effects of day of the week, month, and year on median ride duration. Changed the angle of x-axis titles for easier visualization. 

In terms of seasonal trends, the median duration is consistently lower across all weekdays in January compared to July. This indicates that bike rental durations tend to be shorter in winter (January) compared to summer (July), which is likely due to the colder weather and/or less favorable biking conditions in January.

In terms of weekly patterns, there is an overall increase in median duration towards the weekend, peaking on Saturday and either plateau or dip on Sunday. This pattern is observable in all four plots. In January, there is a slight increase in median ride duration but not too significant, with the increase being more observable in 2020 than 2024. In July, the median duration is higher compared to January, with a more pronounced increase towards the weekend, peaking on Saturday in both 2020 and 2024. There is a sharper increase in median ride duration during the weekend in July 2020 compared to July 2024.

Overall, the plot highlights seasonal and weekly variations in bike rental behavior, with longer rental durations in summer and on weekends.

```{r}
citi_df |> 
  group_by(weekdays, month, year) |> 
  summarize(
    median_duration = median(duration, na.rm = TRUE),
    .group = "drop"
  ) |> 
  mutate(
    weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
  ) |> 
  ggplot(aes(x = weekdays, y = median_duration, group = year, color = month)) +
  geom_line(size = 1) +  
  facet_grid(year ~ month) +
  labs(
    title = "Median Bike Rental Duration by Weekday across January and July of 2020 and 2024", 
    x = "Weekdays", 
    y = "Median Duration",
    caption = "Median Bike Rental Duration by Weekday across January and July of 2020 and 2024") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

Made a violin plot that shows the impact of month, membership status, and bike type on the distribution of ride duration. The plot shows the distribution of ride durations for different bike types, membership statuses (casual vs. member), and months (January and July). 

Overall, casual riders generally have a wider distribution of ride durations compared to members, which suggests that bike members tend to rent for shorter trips while casual riders tend to rent for varying durations. This trend is consistent for both electric and classic bikes, and across the four months. 

For casual riders, classic bikes have a broader range of ride durations, while electric bikes show a higher density around shorter rides. This could suggest that electric bikes are used more frequently for shorter, convenience trips, or it could be because there are not as many electric bikes as there are classic bikes. T

For members, the distributions of ride durations stratified by bike type are relatively similar, with there being a slightly wider range of duration for classic bikes than electric bikes. This difference is not as observable compared to that for casual members. 

In terms of seasonal variation (month and year), there is a slightly higher density of longer rides in July compared to January, particularly for casual riders. This is likely due to better weather conditions. In January, ride durations for both bike types and membership statuses appear shorter, which may be influenced by colder weather. Furthermore, the distribution of ride durations for electric bikes has a wider range in 2024 compared to 2020, possibly due to there being more electric bikes in 2024. This is observable for both casual and members, and across the two seasons. 


```{r}
citi_df |> 
  ggplot(aes(x = rideable_type, y = duration)) +
  geom_violin() +
  facet_grid(month ~ member_casual) +
  labs(
    title = "Distribution of Ride Duration Based on Bike Type, Membership Status and Month",
    x = "Type of Bike",
    y = "Ride Duration (minutes)"
  ) +
  theme_minimal()

```


